---
title: "Unsupervised learning - clustering and dimension reduction"
author: "Anna Yeaton"
date: "Fall 2019"
output:
  html_document:
    df_print: paged
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = T)
```

# Lab Section
Download auto data from the *Statistical Learning* book website here: http://www-bcf.usc.edu/~gareth/ISL/data.html

Today, we are going over Hierarchical clustering, K-Means Clustering, PCA, and ICA. 

```{r load, include=FALSE}
library(ggplot2)
library(tidyverse)
library(ggfortify)
library(fastICA)
```

```{r}
# read in Auto data
Auto_data <- read_csv("Auto.csv")

#remove cars with unknown horsepower and set horsepower to numeric
Auto_data <- Auto_data %>% 
  filter(horsepower != "?") %>% 
  mutate(horsepower = as.numeric(horsepower)) %>% 
  as.data.frame()

#save car names 
Auto_data_names <- Auto_data$name

#data to cluster
Auto_data_clust <- Auto_data[,1:8]
dim(Auto_data_clust)

#392 is too much for a demo, so lets take the first 25
Auto_data_clust <- Auto_data_clust[1:25,]
rownames(Auto_data_clust) <- Auto_data_names[1:25]
```


## Hierarchical agglomerative clustering

Step 1. Assign each item to it's own cluster. We start with 25 clusters, one for each car. 

Step 2. Calculate a proximity matrix between each cluster.

Step 3. Find the pair of clusters closest to each other. 

Step 4. Merge these clusters and then recalculate similarity between clusters. Some options are: single linkage (distance is calculated from the nearest neighbors), complete linkage (distance is calculated from furthest neighbor), average linkage (distance is calculated from mean of different clusters). 

Step 5. Repeat Step 3 and 4 until there is only one cluster.

### In practice

Step 1. Each car is a cluster. 

Step 2. Create a distance matrix from Auto_data_clust.

```{r}
help("dist")
hierarchical_dist <- as.matrix(dist(Auto_data_clust, method = "euclidean"))
#View(hierarchical_dist)
# for k-means you just square this in order to get the euclidean distance squared
```

Step 3. Find the two cars that are the most similar to each other and print the names of those two cars

```{r }
diag(hierarchical_dist) <- NA
arrayInd(which.min(hierarchical_dist), dim(hierarchical_dist))

#postitions 23 and 15 are the most similar. Lets go back to the names of the cars
Auto_data_names[23]
Auto_data_names[15]

```

Step 4. Merge the two clusters together using average linkage. 

```{r }
#replace pos 15 with the average of pos 15 and 23
hierarchical_dist[,15] <- apply((hierarchical_dist[,c(23,15)]),1,mean)
hierarchical_dist[15,] <- apply((hierarchical_dist[c(23,15),]),2,mean)

#remove pos 23
hierarchical_dist <- hierarchical_dist[-23,-23]

#now position 15 represents the cluster containing the saab99e and the toyota corona mark ii
```

Step 5. To complete the algorithm, go back to step 3 and iterate through all of the previous steps until there are no more rows left

```{r }
diag(hierarchical_dist) <- NA #setting all the me vs. me distances to NA
arrayInd(which.min(hierarchical_dist), dim(hierarchical_dist))

#postitions 4 and 3 are the most similar
Auto_data_names[4]
Auto_data_names[3]
# repeat this until we have 1 cluster
```

### R function 

Now that we know how the algorithm works, let's use the R function hclust. Plot the Dendogram resulting from clustering the Auto_data_clust using average linkage.  

```{r}
hierarchical_dist <- dist(Auto_data_clust, method = "euclidean")
tree <- hclust(hierarchical_dist, method="average")
plot(tree)
```

There is one more element to hierarchical clustering: Cutting the tree. Here, we can control how many clusters we want or the height of the tree. 
```{r}
#help(cutree)

# cut tree into 3 clusters
tree <- hclust(hierarchical_dist, method="average")
plot(tree)
tree_k2 <- cutree(tree, k = 2)
# plot the tree before running this line 
rect.hclust(tree, k = 3, h = NULL)
```


\newpage

## Principal Components Analysis (PCA)

Principal Components Analysis is a linear dimensionality reduction algorithm. If you want to learn more about linear algebra, I suggest the MIT Open Courseware class here : https://ocw.mit.edu/courses/mathematics/18-06-linear-algebra-spring-2010/
There are two ways of doing PCA, Single Value Decomposition (SVD), and the method we will use today, using the covariance matrix of the data. 

Step 1. Center data by subtracting the mean.

Step 2. Calculate covariance matrix of data.

Step 3. Perform Eigendecomposition of the covariance matrix. i.e. represent the matrix in terms of it's eigenvalues and eigen vectors

Step 4. Multiply the eigen vectors by the original data to express the data in terms of the eigen vectors. 

Step 1. Center the data by subtracting the mean of the each column from the values in that column


genreally you should always normalize your data unless its already normalized

basically youre changing the way the data is represented in a way that maximizes the variance

```{r}
Auto_data_clust_pca <- data.matrix(Auto_data_clust)

Center_auto <- apply(Auto_data_clust_pca, 2, function(x) x - mean(x))
```


Step 2. Calculate covariance matrix of the Auto data

this will result in a square matrix; by hand can only be done on a square matrix
calculating the variances not the distance
```{r}

Covariance_auto <- cov(Center_auto)
```

Step 3.  Calculate eigen values and vectors

eigenvectors are the axis of the highest variation
eigenvalues are 

```{r}
Eigen_value_auto <- eigen(Covariance_auto)$value

#columns are the eigen vectors
Eigen_vector_auto <- eigen(Covariance_auto)$vector
```

Step 4. Multiply the eigen vector matrix by the original data. 


changes the orientation of the data by multiplying by the eigen vectors

```{r}
PC <- as.data.frame(data.matrix(Center_auto) %*% Eigen_vector_auto)

ggplot(PC, aes(PC[,1], PC[,2])) + geom_point(aes(PC[,1], PC[,2]))
#+ geom_text(aes(label=Auto_data_names[1:8]), nudge_x = -2.5, nudge_y = 400)
```

Step 5. Find out which principal components explain the variance in the data. 

```{r}
#for each component, take the cumulative sum of eigen values up to that point and and divide by the total sum of eigen values
round(cumsum(Eigen_value_auto)/sum(Eigen_value_auto) * 100, digits = 2)
```



Principal component 1 and 2 explain 99.99 percent of the variance. Principal component 1,2, and 3 together explain 100% of the variance in the data. 

### R function 
Now that we know how PCA works, lets use the R funtion prcomp.

```{r}
help("prcomp")
autoplot(prcomp(Auto_data_clust_pca))
```

\newpage

## Independent Component Analysis (ICA)
ICA is an algorithm that finds components that are independent, subcomponents of the data. 

Step 1. Whiten the data by projecting the data onto the eigen vectors (PCA).

Step 2. Solve the X=AS equation by maximizing non-gaussianty in the variables(components) in S. 

This results in a matrix S with components that are independent from each other. 

We will use the fastICA algorithm.

First we will go backwards. 
Create a matrix S with the independent components
```{r}
#create two signals
S <- cbind(cos((1:500)/10), ((500:1)/1000))

par(mfcol = c(1, 2))
plot(S[,1], type="l")
plot(S[,2], type="l")
```

Create a mixing matrix A
```{r}
A <- matrix(c(0.5, 0.7, 0.423, 0.857), 2, 2)
```

Mix S using A


you usually start with the mixed signal and dont know what the original signals are
```{r}
X <- S %*% A
par(mfcol = c(1, 2))
plot(X[,1], type="l")
plot(X[,2], type="l")

```

Unmix using fastICA
```{r, include=FALSE}
a <- fastICA(X, 2, alg.typ = "parallel", fun = "logcosh", alpha = 1,
             method = "R", row.norm = FALSE, maxit = 200,
             tol = 0.0001, verbose = TRUE)
#2 is the number of signals you're looking to get out, 
```


```{r}
par(mfcol = c(1, 2))
plot(1:500, a$S[,1], type = "l", xlab = "S'1", ylab = "")
plot(1:500, a$S[,2], type = "l", xlab = "S'2", ylab = "")
```


### ICA on the auto data

have to be able to pick the right number of signals; seems like guess and check
dont look for more signals than you have samples
```{r, include=FALSE}
a <- fastICA(Auto_data_clust, 7, alg.typ = "parallel", fun = "logcosh", alpha = 1,
             method = "R", row.norm = FALSE, maxit = 200,
             tol = 0.0001, verbose = TRUE)

```

plot the independent components as a heatmap
```{r}
heatmap(a$S)
```




\newpage

## Homework Due 9/19

knit and push both html and rmd

```{r}
data <- iris
View(data)
```

0. Subset the Iris dataset to only include `Sepal.Length`, `Sepal.Width`, `Petal.Length`, and `Petal.Width`. 

```{r}
sub.dat <- data[,1:4]
```

1. Write out the Kmeans algorithm by hand, and run two iterations of it. 
  (dont have to go until all the end)
  
```{r}
#First Iteration
random <- sample(1:3, 150, replace = TRUE)
dat <- cbind(sub.dat, random)
sub1 <- sub.dat[which(dat$random==1),]
sub2 <- sub.dat[which(dat$random==2),]
sub3 <- sub.dat[which(dat$random==3),]
cent1 <- colMeans(sub1)
cent2 <- colMeans(sub2)
cent3 <- colMeans(sub3)
dc1 <- vector()
for (i in 1:nrow(sub.dat)){
  dc1[i] <-(sum((sub.dat[i,] - cent1)^2))
}
dc2 <- vector()
for(i in 1:nrow(sub.dat)){
  dc2[i] <- (sum((sub.dat[i,]-cent2)^2))
}
dc3 <- vector()
for(i in 1:nrow(sub.dat)){
  dc3[i] <- (sum((sub.dat[i,]-cent3)^2))
}
copy <- sub.dat
copy <- cbind(sub.dat, dc1,dc2,dc3)
min.copy <- apply(copy[,5:7], 1, which.min)
copy.new <- cbind(copy,min.copy)

#Second Iteration

sub1.2 <- sub.dat[which(copy.new$min.copy==1),]
sub2.2 <- sub.dat[which(copy.new$min.copy==2),]
sub3.2 <- sub.dat[which(copy.new$min.copy==3),]
cent1.2 <- colMeans(sub1.2)
cent2.2 <- colMeans(sub2.2)
cent3.2 <- colMeans(sub3.2)
dc1.2 <- vector()
for (i in 1:nrow(sub.dat)){
  dc1.2[i] <-(sum((sub.dat[i,] - cent1.2)^2))
}
dc2.2 <- vector()
for(i in 1:nrow(sub.dat)){
  dc2.2[i] <- (sum((sub.dat[i,]-cent2.2)^2))
}
dc3.2 <- vector()
for(i in 1:nrow(sub.dat)){
  dc3.2[i] <- (sum((sub.dat[i,]-cent3.2)^2))
}
sec.it <- sub.dat
sec.it <- cbind(sub.dat, dc1.2,dc2.2,dc3.2)
min.sec.it <- apply(sec.it[,5:7], 1, which.min)
```

2. Run PCA on the Iris dataset. Plot a scatter plot of PC1 vs PC2 and include the percent variance those PCs describe. 


```{r}
PCA <- sub.dat
center_PCA <- apply(PCA, 2, function(x) x - mean(x))
cov_PCA <- cov(center_PCA)
Eval_PCA <- eigen(cov_PCA)$value
Evect_PCA <- eigen(cov_PCA)$vector
pca <- as.data.frame(data.matrix(center_PCA) %*% Evect_PCA)
pca.plot <- ggplot(pca, aes(pca[,1], pca[,2])) + geom_point(aes(pca[,1], pca[,2]))
round(cumsum(Eval_PCA)/sum(Eval_PCA) * 100, digits = 2)

autoplot(prcomp(PCA)) #plot containing percent variance
```

3. Run ICA on the Iris dataset. Plot the independent components as a heatmap.

```{r}
ica <- fastICA(PCA,3, alg.typ = "parallel", fun = "logcosh", alpha = 1,
               method = "R", row.norm = F, maxit = 200,
               tol = 0.0001, verbose = T)

heatmap(ica$S)
```

4. Use Kmeans to cluster the Iris data. 
  * Use the silhouette function in the cluster package to find the optimal number of clusters for kmeans for the iris dataset. Then cluster using kmeans clustering. Does the data cluster by species? 
  * Using this clustering, color the PCA plot according to the clusters.
  
```{r}
k.clust <- kmeans(sub.dat, 3)
dist <- dist(sub.dat)^2
plot(silhouette(k.clust$cluster, dist(sub.dat)))
avg_sil <- function(k){
  k.clust2 <- kmeans(sub.dat, centers = k, nstart = 15)
  sil <- silhouette(k.clust2$cluster, dist(sub.dat))
  mean(sil[,3])
}
k.values <- 2:15
avg_sil_values <- map_dbl(k.values,avg_sil)
plot(k.values, avg_sil_values)

kclust <- kmeans(sub.dat, 2)
plot(silhouette(kclust$cluster, dist(sub.dat)))

#there are three species of flower in the iris dataset, however, when clustered by 3, it shows that the two one group has a stronger clustering while the other two groups cluster weakily. The best cluster number for this data is 2. 

pca.plot <- ggplot(pca, aes(pca[,1], pca[,2])) + geom_point(aes(pca[,1], pca[,2]))

autoplot(prcomp(PCA), data = kclust, colour = kclust$cluster)
autoplot(prcomp(PCA), data = k.clust, colour = k.clust$cluster)
```
  
5. Use hierarchical clustering to cluster the Iris data.

  * Try two different linkage types, and two different distance metrics. 
  * For one linkage type and one distance metric, try two different cut points (cutree function). 
  * Using this clustering, color the PCA plot according to the clusters. (6  plots in total)
  
```{r}
hi.dist <- dist(iris, method = "euclidean")
hi.dist2 <- dist(iris, method = "canberra")

tree1.1 <- hclust(hi.dist, method = "single")
tree1.2 <- hclust(hi.dist, method = "centroid")
tree2.1 <- hclust(hi.dist2, method = "single")
tree2.2 <- hclust(hi.dist2, method = "centroid")

par(mfrow = c(2,2))
plot(tree1.1, hang = -.5)
plot(tree1.2, hang = -.5)
plot(tree2.1, hang = -.5)
plot(tree2.2, hang = -.5)

#For one linkage type and one distance metric, try two different cut points (cutree function)
par(mfrow = c(2,2))
cut2.1 <- cutree(tree2.1, k=3)
plot(tree2.1)
rect.hclust(tree2.1, k = 3, h = NULL)
cut2.2 <- cutree(tree2.1, k = 2)
plot(tree2.1)
rect.hclust(tree2.1, k = 2, h = NULL)


#Using this clustering, color the PCA plot according to the clusters. (6  plots in total) 
#same pca's different colors

autoplot(prcomp(PCA), data = iris, colour = cutree(tree1.1, k=3))
autoplot(prcomp(PCA), data = iris, colour = cutree(tree1.2, k=5))
autoplot(prcomp(PCA), data = iris, colour = cutree(tree2.1, k=2))
autoplot(prcomp(PCA), data = iris, colour = cutree(tree2.1, k=3))
autoplot(prcomp(PCA), data = iris, colour = cutree(tree2.2, k=4))
autoplot(prcomp(PCA), data = iris, colour = cutree(tree2.2, k=6))

````

# Optional material
On PCA:

Eigen Vectors and Eigen Values http://www.visiondummy.com/2014/03/eigenvalues-eigenvectors/
Linear Algebra by Prof. Gilbert Strang https://ocw.mit.edu/courses/mathematics/18-06-linear-algebra-spring-2010/video-lectures/
http://www.cs.otago.ac.nz/cosc453/student_tutorials/principal_components.pdf
https://stats.stackexchange.com/questions/2691/making-sense-of-principal-component-analysis-eigenvectors-eigenvalues

On ICA: 

Independent Component Analysis: Algorithms and Applications https://www.cs.helsinki.fi/u/ahyvarin/papers/NN00new.pdf
Tutorial on ICA taken from http://rstudio-pubs-static.s3.amazonaws.com/93614_be30df613b2a4707b3e5a1a62f631d19.html



